---
title: "DESeq2"
author: "Marco Tello"
date: "2025-05-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# DEA
library(DESeq2)
# Data format and ID conversion
library(data.table)
library(tidyverse)
# Plot generation 
library(ggplot2)
library(factoextra)
library(pheatmap)
# Biological annotation
library(biomaRt)

# Custom functions
custom_volcano <- function(DEG_results, alpha, log2FC){
  # DESeq2 will make some adjusted pvalues NA, details in their docs
  num_na <- sum(is.na(DEG_results$padj))
  # Ommit na results and categorize each gene
  temp <- as.data.table(na.omit(DEG_results))
  temp[, padj := ifelse(padj == 0, 8e-305, padj)]
  temp[, alpha := ifelse(padj < alpha, TRUE, FALSE)]
  temp[, log2FC := ifelse(abs(log2FoldChange) >= log2FC, TRUE, FALSE)]
  temp[, DEG := "No"]
  temp[, DEG := ifelse(alpha & !log2FC, "FDR", DEG)]
  temp[, DEG := ifelse(alpha & log2FC, "FDR and FC", DEG)]
  # Include gene names for the highest fold changes
  # Genes not selected will have their name set to "" in the plot
  setorder(temp, "log2FoldChange")
  # Make plot
  ggplot(temp, aes(x = log2FoldChange, 
                   y = -log10(padj), 
                   color = DEG)) +
    geom_hline(yintercept = -log10(alpha), 
               linetype = "dashed", color = "gray", linewidth = 1) +
    geom_vline(xintercept = log2FC, 
               linetype = "dashed", color = "gray", linewidth = 1) +
    geom_vline(xintercept = -log2FC, 
               linetype = "dashed", color = "gray", linewidth = 1) +
    geom_point(alpha = 0.4, size = 3) +
    scale_x_continuous(breaks = seq(ceiling(range(temp$log2FoldChange))[1],
                                    ceiling(range(temp$log2FoldChange))[2])) +
    scale_color_manual(values = c("No" = "darkgray", 
                                  "FDR" = "blue", 
                                  "FDR and FC" = "red"), 
                       labels = c(paste("FDR (", 
                                        as.character(table(temp$DEG)[1]), 
                                        ")", sep = ""), 
                                  paste("FDR and FC (",
                                        as.character(table(temp$DEG)[2]),")", sep = ""),
                                  paste("Not Sig. (", 
                                        as.character(table(temp$DEG)[3] + num_na), ")", sep = ""))) +
    ylab(expression(-log[10]("FDR")))+
    xlab(expression(log[2]("Fold Change")))+
    labs(color = "DEG status") +
    theme_bw()
}

visualize_degs <- function(DEG_results, alpha, log2FC){
  histogram <- ggplot(na.omit(DEG_results), aes(x = pvalue)) + 
    geom_histogram(binwidth = 0.05, breaks = seq(0, 1, by = 0.05), 
                   fill = "gray", color = "black") + 
    xlab(expression(italic("p-value")))+
    labs(title = "Histogram of p-values", y = "Frequency")+
    theme_bw()
  volcano <- custom_volcano(DEG_results, alpha, log2FC)
  return(list(histogram, volcano))
}


data_path <- "~/GitHub/RNAseq_ribodepleted/output/counts/biotypes"
out_path <- "~/GitHub/RNAseq_ribodepleted/output/deseq2"
  
theme_set(theme_bw())

```


# Differential expression analysis

This report is to identify differentially expressed genes across the following conditions:

1. low glucose control (LG) 
2. high glucose control (HG)
3. high glucose high fat control (HF)
4. high glucose high fat PTS (PTS)
5. high glucose high fat QUE (QUE)


## Format condition names

The first step is to load the selected expression data and format the condition names to match the experimental conditions.

```{r}
# Read exp table
input_file <- file.path(data_path, "protein_coding_exp_filtered-S10-R6.tsv")
exp_df <- fread(input = input_file, drop = c("P2_S2", "Q3_S1"))


# Connect to the Ensembl mart
ensembl <- useEnsembl(
  biomart    = "genes",
  dataset    = "hsapiens_gene_ensembl")

# Query for gene symbol
bm <- getBM(
    attributes = c("ensembl_gene_id", "external_gene_name"),
    filters    = "ensembl_gene_id",
    values     = exp_df$gene_id,
    mart       = ensembl)
exp_df <- merge.data.table(x = bm, y = exp_df, 
                           by.x = "ensembl_gene_id", by.y = "gene_id", 
                           all.x = FALSE, all.y = TRUE)
exp_df <- data.table(exp_df)

# Remove genes with no gene symbol
exp_df <- exp_df[external_gene_name != ""]

# Only one gene symbol has two ENSEMBL IDs so we handle it separately
exp_df[external_gene_name == "POLR2J3", external_gene_name := paste(external_gene_name, 
                                                                    ensembl_gene_id, 
                                                                    sep = "_")]

# Convert data.table to integer matrix
exp_matrix <- exp_df %>%
  dplyr::select(!ensembl_gene_id) %>%
  dplyr::select(!external_gene_name) %>%
  as.matrix()
rownames(exp_matrix) <- exp_df$external_gene_name


# Capture metadata in data frame
condition <- colnames(exp_matrix) %>% 
  str_replace(pattern = "\\d_S\\d+", 
              replacement = "")

# Format condition to match labels
condition <- case_when(
    condition == "C" ~ "HG",
    condition == "F" ~ "HF",
    condition == "L" ~ "LG",
    condition == "P" ~ "PTS",
    condition == "Q" ~ "QUE")
# Create variables for DE analysis
annot <- data.frame(condition = factor(condition, levels = c("LG", "HG", "HF", "PTS", "QUE")),
                    Hglucose  = factor(ifelse(condition == "LG", 0, 1)),
                    Hfat      = factor(ifelse(condition %in% c("LG", "HG"), 0, 1)),
                    PTS       = factor(ifelse(condition == "PTS", 1, 0)),
                    QUE       = factor(ifelse(condition == "QUE", 1, 0)),
                    row.names = colnames(exp_matrix))

annot 

```

Based on the experimental groups we created the appropriate labels and encoded conditions as binary columns to facilitate building models. 

## Set-up DEA parameters

For consistency, we'll define the parameters for differential expression here:

```{r}
alpha <- 0.01
log2FC <- 1
```

The comparisons we'll be looking into are the following:


1. LG vs HG - to see how increasing glucose levels affects gene expression without fat
2. HG vs HF - to isolate the impact of the high fat component under hyperglycemic conditions
3. HF vs PTS -  whether PTS can reverse or alter HF-induced gene changes
4. HF vs QUE - whether QUE can reverse or alter HF-induced gene changes
5. PTS vs QUE - to identify differences in molecular responses between the two compounds.
6. LG vs PTS - to assess whether PTS brings HF-exposed cells back toward the low-glucose baseline
7. LG vs QUE - to assess whether QUE brings HF-exposed cells back toward the low-glucose baseline



Thus, we'll create a section for each of the target comparisons containing:

* Model fitting and result selection
* p-value histograms
* volcano plot
* Output of DEGs


##1. LG vs HG

how increasing glucose levels affects gene expression without fat?


```{r}

# Re-define base condition
annot$condition <- relevel(annot$condition, ref = "LG")
# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)

# Run DESeq2
dds <- DESeq(dds)

# Print comparisons made
resultsNames(dds)
```


```{r}

comparison <- "condition_HG_vs_LG"

de_res <- results(dds,
                  name=comparison,
                  pAdjustMethod = "BH") 
temp <- as.data.table(de_res)
de_res <- data.table(geneID = row.names(de_res), temp)


out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(de_res, alpha, log2FC)
temp

```


##2. HG vs HF


```{r}

# Re-define base condition
annot$condition <- relevel(annot$condition, ref = "HG")
# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)

# Run DESeq2
dds <- DESeq(dds)

# Print comparisons made
resultsNames(dds)
```


```{r}

comparison <- "condition_HF_vs_HG"

de_res <- results(dds,
                  name=comparison,
                  pAdjustMethod = "BH") 
temp <- as.data.table(de_res)
de_res <- data.table(geneID = row.names(de_res), temp)


out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(de_res, alpha, log2FC)
temp
```

##3. HF vs PTS



```{r}

# Re-define base condition
annot$condition <- relevel(annot$condition, ref = "HF")
# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)

# Run DESeq2
dds <- DESeq(dds)

# Print comparisons made
resultsNames(dds)
```


```{r}
 
comparison <- "condition_PTS_vs_HF"

de_res <- results(dds,
                  name=comparison,
                  pAdjustMethod = "BH") 
temp <- as.data.table(de_res)
de_res <- data.table(geneID = row.names(de_res), temp)


out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(de_res, alpha, log2FC)
temp

```




##4. HF vs QUE



```{r}

# Re-define base condition
annot$condition <- relevel(annot$condition, ref = "HF")
# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)

# Run DESeq2
dds <- DESeq(dds)

# Print comparisons made
resultsNames(dds)
```


```{r}

comparison <- "condition_QUE_vs_HF"

de_res <- results(dds,
                  name=comparison,
                  pAdjustMethod = "BH") 
temp <- as.data.table(de_res)
de_res <- data.table(geneID = row.names(de_res), temp)


out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(de_res, alpha, log2FC)
temp
```




##5. PTS vs QUE


```{r}

# # Make DESeq object
# dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
#                               colData = annot,
#                               design= ~ Hglucose + Hfat + PTS + QUE)
# # Run DESeq2
# dds <- DESeq(dds)
# 
# # Print comparisons made
# resultsNames(dds)
# 
# # Contrast positions:
# # 1. Intercept        = LG baseline
# # 2. Hglucose_1_vs_0  = LG  → HG  (add high glucose)
# # 3. Hfat_1_vs_0      = HG  → HF  (add high fat)
# # 4. PTS_1_vs_0       = HF  → PTS (add PTS)
# # 5. QUE_1_vs_0       = HF  → QUE (add QUE)
# de_res <- results(dds,
#                   contrast = c(0, 0, 0, 1, -1 ), # Reference is PTS (4 element)
#                   pAdjustMethod = "BH") 

```


```{r}

comparison <- "condition_QUE_vs_PTS"

annot$condition <- relevel(annot$condition, ref = "PTS")

# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)


temp <- as.data.table(de_res)
temp <- data.table(geneID = row.names(de_res), temp)

out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(as.data.table(de_res), alpha, log2FC)
temp
```



##6. LG vs PTS


```{r}
comparison <- "condition_PTS_vs_LG"

annot$condition <- relevel(annot$condition, ref = "LG")

# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)


temp <- as.data.table(de_res)
temp <- data.table(geneID = row.names(de_res), temp)

out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(as.data.table(de_res), alpha, log2FC)
temp

```


##7. LG vs QUE


```{r}

comparison <- "condition_QUE_vs_LG"

annot$condition <- relevel(annot$condition, ref = "LG")

# Make DESeq object
dds <- DESeqDataSetFromMatrix(countData = exp_matrix,
                              colData = annot,
                              design= ~ condition)


temp <- as.data.table(de_res)
temp <- data.table(geneID = row.names(de_res), temp)

out_file <- file.path(out_path, paste("deseq2_", 
                                      str_replace_all(comparison, "_", ""),
                                      "_FDR", str_replace(alpha, "\\.", ""),
                                      ".tsv", sep = ""))
if(!file.exists(out_file)){
  output_table = de_res
  output_table[, log2FoldChange := round(log2FoldChange, 4)]
  output_table[, baseMean       := round(baseMean, 4)]
  output_table[, lfcSE          := round(lfcSE, 4)]
  output_table[, stat           := round(stat, 4)]
  output_table[, pvalue         := round(pvalue, 4)]
  output_table[, padj           := round(padj, 4)]
  output_table[, FoldChange := round((2**abs(log2FoldChange))*sign(log2FoldChange), 4)]
  fwrite(x = output_table, file = out_file, 
         append = FALSE, quote = FALSE, sep = '\t', 
         row.names = FALSE, col.names = TRUE)
}

temp <- visualize_degs(as.data.table(de_res), alpha, log2FC)
temp

```




